Continuous Deployment and Schema Evolution \\in SQL Databases

### 2015-02-22 paper.tex
Preparing camera-ready version by listing feedback.
Processed some of the feedback provided by reviewers.

\author{
  \IEEEauthorblockN{
    Michael de Jong\IEEEauthorrefmark{1},
    Arie van Deursen\IEEEauthorrefmark{2}
  }
  \IEEEauthorblockA{
    Delft University of Technology,\\
    Delft, The Netherlands\\ \\
    Email: M.deJong-2@student.tudelft.nl\IEEEauthorrefmark{1}\\
    Email: Arie.vanDeursen@tudelft.nl\IEEEauthorrefmark{2}\\ \\
    \today{}
  }
}

@@

\begin{itemize}
  \item None of the tools used in practice support foreign keys.
  \item The tools primarily focus on MySQL databases.
  \item These tools are limited in the sense that they typically only support making small changes to the existing database schema. This forces software and release engineers to often release intermediate versions of the application, as the schema is altered in several steps.
\end{itemize}

@@

The previously discussed tools from practice all use Approach~1 or a variation of it. They prove that this is a viable way for simple evolutions of the database schema. However where these tools all create a single ghost table, we aim to create one ghost table for each table that is under change, and ghost tables for all tables which in turn are dependent on these tables under change. For instance adding a new non-nullable column ``summary'' to the ``books'' table (see Figure~\ref{fig:migration-stages}), means we should also create a ghost table for the ``publications`` table since there is a foreign key pointing from the ``publications'' table to the ``books'' table. The other tables do not require ghost tables, as they do not have any foreign keys referencing either the ``books'' or ``publications'' table. This approach ensures that we can support foreign keys.

@@

@@

\subsection{Current Status}

@@

\begin{figure}%
    \centering
    \subfloat[Original database schema, before migration]{{\includegraphics[width=3.0in]{pre-migration-state.png} }}%
    \qquad
    \subfloat[Database schema in mixed-state, during migration]{{\includegraphics[width=3.0in]{in-migration-state.png} }}%
    \qquad
    \subfloat[New database schema, after migration]{{\includegraphics[width=3.0in]{post-migration-state.png} }}%
    \caption{Query performance of demo application before and during migration}%
    \label{fig:migration-stages}%
\end{figure}

QuantumDB currently consists of two parts: 
\begin{itemize}
  \item{\textbf{qdb:core} is the program responsible for performing the actual migration tasks on the database. It does this by creating ghost tables of tables under change, using triggers to keep the ghost tables and the original tables synchronized with each other, and a background task which copies the data from the original tables to the ghost tables in small batches to prevent degrading the performance of the database.}
  \item{\textbf{qdb:driver} is a Java JDBC driver which can wrap another JDBC driver. When used, the software engineers can specify a database schema version on which that version of their application operates. The driver then rewrites SQL queries sent to it by the application, and passes them on to the wrapped database JDBC driver which in turn sends them to the database. When rewriting queries, it replaces table names with the names of the associated (ghost) table names where applicable.}
\end{itemize}

@@

The second approach uses QuantumDB to add the new column. It first creates a ghost table based on the ``users'' table with the new column. It then iteratively copies records from the ``users'' table to the ghost table in batches of 2,000 records at intervals of 50 milliseconds. With this we attempt to avoid negatively affecting the performance of the queries issued by the demo application while we are copying the records. Unlike with the previous approach, with this approach the demo applications use the QuantumDB JDBC driver which exposes each demo application to a different version of the database schema (containing either the ``users'' table or the ghost table). The driver ensures that each query is rewritten in such a way that the correct table is queried for that specific version of the schema.

@@

\begin{figure}%
    \centering
    \subfloat[Before migration]{{\includegraphics[width=1.55in]{normal-operations.png} }}%
    \qquad
    \subfloat[During migration]{{\includegraphics[width=1.55in]{migration.png} }}%
    \caption{Query performance of demo application before and during migration}%
    \label{fig:migration}%
\end{figure}

It is worth noting that we have not invested any time in improving the performance of QuantumDB or reducing the performance impact it has on the database and its clients. During this experiment we noticed that the hard disk was periodically unable to write data to disk quickly enough. As a result queries were taking significantly longer than normal to complete. We are confident that we can greatly reduce this by tweaking the PostgreSQL config, switching to a faster hard disk, or lowering the rate of migration during the copy phase.

@@

\begin{itemize}
  \item{QuantumDB can already deal with foreign keys referring from tables under change to other tables not under change. To fully support foreign keys QuantumDB would also need to be able to handle foreign keys referring to tables under change. This is currently being worked on.}
  \item{Use performance metrics to throttle the copy phase. When the database is (too) busy (high disk I/O, high CPU usage, or memory starvation), we should slow down the iterative copy process. Vice-versa we can speed it up when the database is not under load.}
  \item{QuantumDB currently only supports PostgreSQL, but since most SQL databases have support for triggers and procedures of some kind, this should prove easily extendable to other popular SQL databases.}
  \item{The driver currently uses a very rudimentary query rewriting algorithm to replace table names under change with the names of ghost tables. It does not yet support the entire PostgreSQL query syntax.}
\end{itemize}

@@

Although development of QuantumDB is still ongoing, initial measurements are promising. With our small test application we have shown that we can migrate from one database schema to a state where both the old and the new database schemas are active, while still enabling database clients to remain operational. Applications can connect with the database according to a specific schema version, and no longer have to deal with database schemas being in an in-between state. This should effectively allow further decoupling of the release process from the development process. We believe that this approach can put \textit{Continuous Deployment} and \textit{Canary Releases} in reach of more software and release engineers.

>>>>

\author{
  \IEEEauthorblockN{
    Michael de Jong\IEEEauthorrefmark{1},
    Arie van Deursen\IEEEauthorrefmark{2}
  }
  \IEEEauthorblockA{
    Delft University of Technology,
    Delft, The Netherlands\\
    Email: M.deJong-2@student.tudelft.nl\IEEEauthorrefmark{1}
    Email: Arie.vanDeursen@tudelft.nl\IEEEauthorrefmark{2}\\ \\
    \today{}
  }
}

@@

\begin{itemize}
  \item None of the tools used in practice support foreign keys.
  \item The tools discussed primarily focus on MySQL databases. \textit{OpenArk Kit} and \textit{Percona Toolkit} are suits of tools which focus exclusively on MySQL database administration.
  \item These tools are limited in the sense that they typically only support making small changes to the existing database schema. This forces software and release engineers to often release intermediate versions of the application, as the schema is altered in several steps.
\end{itemize}

@@

The previously discussed tools from practice all use Approach~1 or a variation of it. They prove that Approach~1 is a viable way for small evolutions of the database schema. However where these tools all create a single ghost table, we aim to create one ghost table for each table that is under change, and ghost tables for all tables which in turn are dependent on these tables under change. For instance adding a new non-nullable column ``summary'' to the ``books'' table (see Figure~\ref{fig:migration-stages}), means we should also create a ghost table for the ``publications`` table since there is a foreign key pointing from the ``publications'' table to the ``books'' table. The other tables do not require ghost tables, as they do not have any foreign keys referencing either the ``books'' or ``publications'' table.

@@

Another difference worth mentioning is that in our approach the tables never actually switch names. Ultimately when making the switch we simply delete the original table (and its dependencies), and continue to read from and write to the ghost table.

@@

@@

\begin{figure}%
    \centering
    \subfloat[Original database schema]{{\includegraphics[width=2.6in]{pre-migration-state.png} }}%
    \qquad
    \subfloat[Database schema in mixed-state]{{\includegraphics[width=2.6in]{in-migration-state.png} }}%
    \qquad
    \subfloat[New database schema]{{\includegraphics[width=2.6in]{post-migration-state.png} }}%
    \caption{Different stages of the database schema when modifying the ``books'' table.}%
    \label{fig:migration-stages}%
\end{figure}

QuantumDB currently consists of two parts: 
\begin{itemize}
  \item{\textbf{qdb:core} is the program responsible for performing the actual migration tasks on the database. It does this by creating ghost tables of tables under change, using triggers to keep the ghost tables and the original tables synchronized with each other, and a background task which copies the data from the original tables to the ghost tables in small batches to prevent degrading the performance of the database. We know that for PostgreSQL triggers are executed within the same transaction as what triggered the trigger. In our case it means that any modification to the original tables is also performed on the ghost tables in the same transaction.}
  \item{\textbf{qdb:driver} is a Java JDBC driver which can wrap another JDBC driver. When used, the software engineers can specify a database schema version on which that version of their application operates. The driver then rewrites SQL queries sent to it by the application, and passes them on to the wrapped database JDBC driver which in turn sends them to the database. When rewriting queries, it replaces table names with the names of the associated (ghost) table names where applicable.}
\end{itemize}

@@

The second approach uses QuantumDB to add the new column. It first creates a ghost table based on the ``users'' table with the new column. It then iteratively copies records from the ``users'' table to the ghost table in batches of 2,000 records at intervals of 50 milliseconds. With this we attempt to avoid negatively affecting the performance of the queries issued by the demo application while we are copying the records. With this approach the demo applications use the QuantumDB JDBC driver which exposes each demo application to a different version of the database schema. The driver ensures that each query is rewritten in such a way that the correct table is queried for that specific version of the schema.

@@

\begin{figure}%
    \centering
    \subfloat[Before migration]{{\includegraphics[width=1.55in]{normal-operations.png} }}%
    \qquad
    \subfloat[During migration]{{\includegraphics[width=1.55in]{migration.png} }}%
    \caption{Query performance before and during migration}%
    \label{fig:migration}%
\end{figure}

We have not invested any time in improving the performance of QuantumDB or reducing the performance impact it has on the database and its clients. During this experiment we noticed that the hard disk was periodically unable to write data to disk quickly enough. As a result queries were taking significantly longer than normal to complete. We are confident that we can greatly reduce this by tweaking the PostgreSQL config, switching to a faster hard disk, or lowering the rate of migration during the copy phase.

@@

\begin{itemize}
  \item{QuantumDB can already deal with foreign keys referring from tables under change to tables not under change. To fully support foreign keys QuantumDB would also need to be able to handle foreign keys referring to tables under change.}
  \item{Use performance metrics to throttle the copy phase. When the database is (too) busy (high disk I/O, high CPU usage, or memory starvation), we should slow down the iterative copy process. Vice-versa we can speed it up when the database is not under load.}
  \item{QuantumDB currently only supports PostgreSQL, but since most SQL databases have support for triggers and procedures of some kind, this should prove easily extendable to other popular SQL databases.}
  \item{The driver currently uses a very naive method of query rewriting to replace table names under change with the names of ghost tables. The PostgreSQL query syntax is only partially supported.}
\end{itemize}

@@

Development of QuantumDB is still ongoing, but initial measurements are promising. With our small test application we have shown that we can migrate from one database schema to a state where both the old and the new database schemas are active, while still enabling database clients to remain operational. Applications can connect with the database according to a specific schema version, and no longer have to deal with database schemas being in an in-between state. This should effectively allow further decoupling of the release process from the development process. We believe that this approach can put \textit{Continuous Deployment} and \textit{Canary Releases} in reach of more software and release engineers.
